---
output:
  html_document:                    # Classe de documento.
    toc: true                       # Exibir sum?rio.
    toc_depth: 2                    # Profundidade do sum?rio.
    toc_float:                      # Sum?rio flutuante na borda.
      collapsed: true
      smooth_scroll: true
    number_sections: true           # Se??es numeradas.
    theme: united
    #spacelab
    #default,cerulean,journal,flatly,readable,spacelab,
    #united,cosmo,lumen,paper,sandstone,simplex,yeti
    
    highlight: espresso
    #default, tango, pygments, kate, monochrome, espresso, zenburn, haddock, and textmate
    #css: styles.css                 # Caminho para arquivo CSS.
    fig_width: 7                    # Lagura das figuras.
    fig_height: 6                   # Altura das figuras.
    fig_caption: true               # Exibica??o de legenda.
    fig_align: 'center'
    code_folding: hide              # Esconder/exibir bloco de c?digo.
#    keep_md: true                   # Manter o arquivo md.
    #template: quarterly_report.html # Caminho para o template.  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE,fig.align='center',
                      fig.height = 5, fig.width = 6)
```

---


<center>
<font size="6"> 
<p align=”center”> <b> Testes de hipótese </b> </center>
</font>
</center>

---

  * Análises de variância.
  * Hipóteses lineares.
  * Erros correlacionados.
  * Variância não constante.
  * Outliers. 
  * etc.

---

# Funções

```{r, echo=FALSE, message=FALSE}
library(car)

funcoes <- read.csv2("class_func.csv", header = T, sep = ',')

tabela <-funcoes[, c(2,5,6)]
names(tabela) <- c('Função', 'Categoria', 'Utilidade')

DT::datatable(subset(tabela, Categoria == "TESTES DE HIPÓTESE"), 
              filter = 'top',
              options = list(
                bPaginate = FALSE))
               

```

---

# Exemplos

---

## Teste Levene {.tabset .tabset-fade}

Computes Levene's test for homogeneity of variance across groups.

---

### Exemplo 1

```{r}
with(Moore, leveneTest(conformity, fcategory))
with(Moore, leveneTest(conformity, interaction(fcategory, partner.status)))
leveneTest(conformity ~ fcategory*partner.status, data=Moore)
```

---

### Exemplo 2

```{r}
leveneTest(lm(conformity ~ fcategory*partner.status, data=Moore))
leveneTest(conformity ~ fcategory*partner.status, data=Moore, center=mean)
leveneTest(conformity ~ fcategory*partner.status, data=Moore, center=mean, trim=0.1)
```

---

## Teste Durbin-Watson

Durbin-Watson Test for Autocorrelated Errors

```{r}
m <- lm(fconvict ~ tfr + partic + degrees + mconvict, data=Hartnagel)
durbinWatsonTest(m)
plot(m,3)
```

---

## Teste Score

Score Test for Non-Constant Error Variance

Computes a score test of the hypothesis of constant error variance against the alternative that the error variance changes with the level of the response (fitted values), or with a linear combination of predictors

```{r}
ncvTest(lm(interlocks ~ assets + sector + nation, data=Ornstein))

ncvTest(lm(interlocks ~ assets + sector + nation, data=Ornstein), 
        ~ assets + sector + nation, data=Ornstein)
```

---

## Teste Bonferroni

Bonferroni Outlier Test

Reports the Bonferroni p-values for testing each observation in turn to be a mean-shift outlier, based Studentized residuals in linear (t-tests), generalized linear models (normal tests), and linear mixed models.

```{r}
m <- lm(prestige ~ income + education, data=Duncan)
outlierTest(m)
```

---

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>